---
title: "Class 9: Candy Mini-Project"
author: "Rowen PID: A18363990"
format: pdf
editor: visual
toc: true
---

## Background

In this mini-project, you will explore FiveThirtyEight’s Halloween Candy dataset. FiveThirtyEight, sometimes rendered as just 538, is an American website that focuses mostly on opinion poll analysis, politics, economics, and sports blogging. They recently ran a rather large poll to determine which candy their readers like best.

## Importing Candy Data

```{r}
candy_file <- "candy-data.csv"

candy <- read.csv("candy data/candy-data.csv", row.names = 1)

head(candy)
```

> > Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> > Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

```{r}
candy["Twix", ]$winpercent

library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```

> > Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["Kit Kat", ]$winpercent
```

My favorite candy is Kit Kat and its winpercent is 76.7686

> > Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> > Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

> > Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the `winpercent` variable is on a different scale. Most other variables are between 0 and 1, while `winpercent` ranges from 0 to 100.

> > Q7. What do you think a zero and one represent for the candy\$chocolate column?

A value of 1 means the candy contains chocolate, while 0 means it does not.

## Exploratory analysis

> > Q8. Plot a histogram of `winpercent` values using both base R and ggplot2.

```{r}
hist(candy$winpercent,
     main = "Winpercent Distribution",
     xlab = "Winpercent",
     col = "lightblue")

library(ggplot2)

ggplot(candy, aes(x = winpercent)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(title = "Winpercent Distribution")
```

> > Q9. Is the distribution of `winpercent` values symmetrical?

The distribution of `winpercent` values is not perfectly symmetrical and it is slightly skewed.

> > Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

The center of the distribution is slightly above 50%.

> > Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[candy$chocolate == 1])
mean(candy$winpercent[candy$fruity == 1])
```

On average, chocolate candies have a higher winpercent than fruity candies.

> > Q12. Is this difference statistically significant?

```{r}
t.test(
  candy$winpercent[candy$chocolate == 1],
  candy$winpercent[candy$fruity == 1]
)
```

Yes, the difference is statistically significant because the p-value (2.871e-08) is much less than 0.05. This shows that there is a significant difference in `winpercent` between chocolate and fruity candies.

## Overall Candy Rankings

> > Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent), ], 5)
```

The 5 least liked candy types are: Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

> > Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(-candy$winpercent), ], 5)
```

The top 5 favorite candy types are: Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kit, and Snickers.

> > Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col() + 
  theme(axis.text.y = element_text(size = 3))
```

> > Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col() +
  theme(axis.text.y = element_text(size = 3))
```

```{r}
my_cols = rep("black", nrow(candy))

my_cols[as.logical(candy$chocolate)] = "chocolate"

my_cols[as.logical(candy$bar)] = "brown"

my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill = my_cols) +
  ylab("") +
  theme(axis.text.y = element_text(size = 3))
```

> > Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate is Sixlets.

> > Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starbust.

## Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(winpercent, pricepercent, label = rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

>> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Peanut Butter Cup

>> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
The five most expensive candies are Nik L Nip, Nestle Smarties, Ring Pop, Hershey’s Krackel, and Hershey’s Milk Chocolate. Among these, Nik L Nip is the least popular because it has the lowest winpercent.

## Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

>> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are anti-correlated because candies tend to be either chocolate based or fruit flavored, but not both.

>> Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

I expect chocolate and fruity to have the largest contributions to PC1 because they are anti-correlated and show a big difference between types of candy, which gives separation along the first principal component.

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1], pca$x[,2])
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) +
  aes(
    x = PC1,
    y = PC2,
    size = winpercent/100,
    text = rownames(my_data),
    label = rownames(my_data)
  ) +
  geom_point(col = my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
ggplot(pca$rotation) +
  aes(
    x = PC1,
    y = reorder(rownames(pca$rotation), PC1)
  ) +
  geom_col()
```

>> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

PC1 has strong positive loadings for fruity, pluribus, and hard, representing fruity multi piece candies. This makes sense and reflects the earlier observed contrast between fruity and chocolate candies.

## Summary

>> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

Winning candies are usually chocolate based, bar type candies with ingredients like caramel or nuts. Visualizations showed higher rankings for these candies, correlation showed chocolate and fruity are opposites, and PCA confirmed this by separating them along PC1. All methods point to the same pattern.








