---
title: "class12"
author: "Rowen Lin (PID: A18363990)"
format: pdf
editor: visual
toc: true
---

## Section 4: Population Scale Analysis

>> Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt", header=TRUE)

head(expr)

table(expr$geno)

by(expr$exp, expr$geno, median)
```
The dataset contains 462 total samples. Sample sizes for each genotype: A/A = 108, A/G = 233, G/G = 121. The median expression levels are: A/A = 31.25, A/G = 25.06, and G/G = 20.07. This shows a clear trend where having more G alleles is associated with lower ORMDL3 expression.

>> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(expr, aes(x=geno, y=exp, fill=geno)) +
  geom_boxplot(notch=TRUE) +
  geom_jitter(alpha=0.3, width=0.2) +
  labs(x="Genotype", y="Expression",
       title="ORMDL3 Expression by Genotype") +
  theme_bw() +
  scale_fill_manual(values=c("red","green","blue")) +
  theme(legend.position="none")
```

