---
title: "Class07"
author: "Rowen (PID: A18363990)"
format: pdf
editor: visual
---

# First up kmeans()

Demo of using kmeans() function in base R. First make up some data with a known structure.

```{r}
tmp <- c( rnorm(30, -3), rnorm(30, 3) )
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```

Now we have some made up data in `x` let's see how kmeans works with this data

```{r}
k <- kmeans(x, centers = 2, nstart = 20)
k
```

> Q. How many points are in each cluster

```{r}
k$size
```

> Q. How we do get to the cluster membership/assignment.

```{r}
k$cluster
```

> Q. What about cluster centers?

```{r}
k$centers
```

Now we got to the main results let's use them to plot our data with the kmeans result

```{r}
plot(x, col=k$cluster)
points(k$centers, col="blue", pch=15)
```

## Now for Hierarchical Clustering

We will cluster some data `x` with the `hclust()`. In this case `hclust()` requires a distance matrix as input.

```{r}
hc <- hclust( dist(x) )
hc
```

Let's plot our hclust result

```{r}
plot(hc)
```

To get our cluter membership vector we need to "cut" the tree with the `cutree()`

```{r}
grps <- cutree(hc, h=8)
grps
```

Now plot our data with the hclust() results.

```{r}
plot(x, col=grps)
```

# Principal Component Analysis (PCA)

## PCA of UK food data

Read data from website and try a few visualizations.

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
x
```

> > **Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?**

```{r}
dim(x)
```

x has 17 rows and 5 columns.

> > **Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?**

I prefer storing the row names as a column instead of using `x <- x[, -1]`. This is more robust because if you run `x <- x[, -1]` multiple times, it will keep deleting columns and can remove actual data by mistake. Keeping the names as a column is safer and works better for plotting and analysis.

> > **Q3. Changing what optional argument in the above barplot() function results in the following plot?**

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

```{r}
barplot(as.matrix(x), beside = FALSE, col = rainbow(nrow(x)))
```

```{r}
library(tidyr)

# Convert data to long format for ggplot with `pivot_longer()`
x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)
```

> > **Q4. Changing what optional argument in the above ggplot() code results in a stacked barplot figure?**

```{r}
library(ggplot2)

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_bw()
```

```{r}
library(ggplot2)

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "stack") +
  theme_bw()
```

> > **Q5. We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?**

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

The `pairs()` function makes scatterplots for every pair of variables, so we can compare countries. Each point is a food category. If a point is on the diagonal, it means the two countries have the same values for that food. Points far from the diagonal mean they are different.

> > **Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?**

```{r}
library(pheatmap)
```

England, Scotland, and Wales cluster together, while N. Ireland is more separate. This suggests those three countries have similar food consumption patterns, but N. Ireland is different.

It’s not super easy to see exactly which foods are different, but the heatmap shows N. Ireland has different levels for some categories compared to the others.

PCA to the rescue!! The main base R PCA function is called `prcomp()` and we will need to give it the transpose of our input data!

```{r}
pca <- prcomp(t(x))
```

There is a nice summary of how well PCA is doing

```{r}
summary(pca)
```

```{r}
attributes(pca)
```

```{r}
pca <- prcomp(t(x), scale. = TRUE)
```

To make our new PCA plot (a.k.a. PCA score plot) we access `pca$x`

```{r}
plot(pca$x[,1], pca$x[,2])
text(pca$x[,1], pca$x[,2], colnames(x))
```

color up the plot

```{r}
country_cols <- c("orange", "red", "blue", "green")
plot(pca$x[,1], pca$x[,2])
text(pca$x[,1], pca$x[,2], colnames (x))

```

> > **Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.**

```{r}
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

ggplot(df) +
  aes(x = PC1, y = PC2, label = Country) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

> > **Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.**

```{r}
ggplot(df) +
  aes(x = PC1, y = PC2, label = Country, color = Country) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  scale_color_manual(values = c(
    "England" = "orange",
    "Scotland" = "blue",
    "Wales" = "red",
    "N.Ireland" = "darkgreen"
  )) +
  theme_bw()
```

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v

z <- summary(pca)
z$importance
```

```{r}
variance_df <- data.frame(
  PC = factor(paste0("PC", 1:length(v)), levels = paste0("PC", 1:length(v))),
  Variance = v
)

ggplot(variance_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0))
```

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

> > **Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?**

```{r}
ggplot(pca$rotation) +
  aes(x = PC2,
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

PC2 is mainly influenced by foods like Fresh_potatoes and Soft_drinks, which have the largest loading values.

This means PC2 separates countries based on differences in these types of foods. Countries with higher values eat more of these foods, while others eat less.

## PCA of RNA-Seq data

Read in data from website

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```

```{r}
pca <- prcomp( t(rna.data))
summary(pca)
```

Do our PCA plot of this RNA

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
text(pca$x[,1], pca$x[,2], colnames(rna.data))
```
