---
title: "Class5: Data Viz with ggplot2"
author: "Rowen (PID: A18363990)"
date: "April 14, 2026"
format: pdf
editor: visual
---

#Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **gglplot2**.

Let's compare how we make a simple figure with these two systems:

We can use the in-built 'cars' dataset:

```{r}
plot(cars)
```

Before I can use ggplot 2 I need to install it on my computer.

To do this we can use the function `install.packages("ggplot2")`

\>**N.B.** We never run `install.packages()`in our quarto doc (we run it once only in our R console) as it would re-install the package every time we render our report.

Once installed we need to load up the package into our R brain:

```{r}
library(ggplot2)
```

The main function in the ggplot2 package is called `ggplot()`

```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers:

-   The **data** (a data.frame of the stuff we want to plot)

-   The **aes**thetics (how the data maps to the plot)

-   The **geom layer (how you want the plot drawn, e.g. points, lines, etc.)**

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

## Add some custom features

Let's add a trend line that shows the relationship between speed and distance.

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_line()
```

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  theme_bw() + 
  labs(title="Stoping Dist of old cars",
       x="Speed (MPH)",
       y="Distance (ft)")
```

Q. Can you make the `geom_smooth()` function produce a linear straight line fit to the data and turn off the "gray" error region.

Important data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
sum(genes$State == "up")
```

A useful new function in this context is the `table()` function:

```{r}
table(genes$State)
```

My first plot attempt

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +  
  geom_point() +
  scale_color_manual(values=c("purple", "gray", "orange")) +
  theme_bw()+
  labs(x="No drug", y="Drug",
       title="Expression changes upon GPL-1 inhibitor treatment")
```

## Going further

Here we read the famous gapminder data set:

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many entries (i.e. rows) are in this data set?

```{r}
nrow(gapminder)
```

> Q. How many different "country" entries are in this dataset?

```{r}
length(unique(gapminder$country))
```

```{r}
unique(gapminder$country)
```

Let's make our first plot of the entire dataset:

Plot of "gdpPercap" vs "lifeExp" colored by "continent"

```{r}
p <- ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=0.3)
```

```{r}
p
```

I can add more layers to \`p\`

```{r}
p +
  facet_wrap(~continent)
```

Make a plot for 1977 and 2007 only (not all the years in the data set).

> Q. First use the **dplyr** package and the `filter()` function from that package to extract the year 2007.

```{r}
library (dplyr)
```

```{r}
g07 <- filter(gapminder, year ==2007)
g77 <- filter(gapminder, year ==1977)
g <- filter(gapminder, year ==2007 | year ==1977)
```

```{r}
ggplot(g07) +
  aes(gdpPercap, lifeExp, col=continent, size=pop) +
  geom_point() +
  facet_wrap(~year)
```

> Q, Make a histogram of lifeExp colored by continent (try using `fill=continent` or `col=continent`)
>
> Q. Make a histogram of lifeExp faceted by continent

```{r}
ggplot(gapminder) +
  aes(x = lifeExp, col = continent) +
  geom_histogram()
```
