---
title: "Class 11 Protein Structure Prediction with AlphaFold"
author: "Rowen Lin (PID: A18363990)"
format: pdf
editor: visual
toc: true
---

```{r}
library(bio3d)
```

```{r}
unzip("hipvrdimer_94b5b.result.zip", exdir = "hivprdimer")

results_dir <- "hivprdimer/hivprdimer_94b5b/"
```


Make a vector of input PDB file names that we can read into R

```{r}
pdb_files <- list.files(path = results_dir, pattern = "*.pdb", full.names = TRUE)
pdb_files
```

## Custom Analysis of AlphaFold Models

```{r}
pdbs <- pdbaln(pdb_files, fit = TRUE, exefile = "msa")
pdbs
```

Here we read in all 5 AlphaFold model PDB files and superposed them onto each other using the `pdbaln()` function from bio3d.

```{r}
rd <- rmsd(pdbs)
print(rd, digits=2)
```

```{r}
pdb <- read.pdb(pdb_files[1])

plotb3(pdb$atom$b[pdb$calpha], typ="l", lwd=2, sse=pdb,
       ylab="B-factor (pLDDT)")
```
The plot shows the pLDDT confidence scores per residue for the top 
ranked model. Values above 70 are considered high confidence. The 
HIV-Pr monomer shows consistently high pLDDT scores across the entire 
chain, indicating a reliable prediction throughout.

## Predicted Alignment Error for domains

>> Next we examine the Predicted Aligned Error (PAE) matrix, which tells 
us about the confidence in the relative positions of residue pairs 
across the structure.

```{r}
library(jsonlite)

pae_file <- list.files(results_dir, pattern="predicted_aligned_error_v1.json",
                       full.names=TRUE)
pae <- fromJSON(pae_file)

plot.dmat(pae$predicted_aligned_error, xlab="Residue", ylab="Residue",
          main="Predicted Aligned Error (PAE)")
```
The PAE matrix shows mostly blue across the entire residue matrix, which shows that AlphaFold is confident in the relative positions of all residue pairs. 

## Residue conservation from alignment file

```{r}
aln_file <- list.files(results_dir, pattern=".a3m", full.names=TRUE)
aln <- read.fasta(aln_file[1], to.upper=TRUE)
dim(aln$ali)
```
The alignment has 5397 sequences across 132 positions.

```{r}
sim <- conserv(aln)

plotb3(sim[1:99], sse=pdb, ylab="Conservation Score")
```

```{r}
con <- consensus(aln, cutoff=0.9)
con$seq

pdb$atom$b[pdb$calpha] <- sim[1:99]
write.pdb(pdb, file="m1_conserv.pdb")
```

## Section 10: Summary

In this lab I used ColabFold to generate AlphaFold2 structure predictions for HIV protease. I analyzed the results using bio3d in R, and then I found pLDDT confidence scores, RMSD between models, predicted aligned error, and residue conservation. All 5 of these models showed high confidence and were very similar structures, with the catalytic Asp25 identified as the most conserved residue.






