---
title: "Class 13: Transcriptomics and the analysis of RNA-Seq data"
author: "Rowen Lin (PID: A18363990)"
format: pdf
editor: visual
toc: true
---

## Background

The data for this hands-on session comes from a published RNA-seq experiment where airway smooth muscle cells were treated with dexamethasone, a synthetic glucocorticoid steroid with anti-inflammatory effects (Himes et al. 2014).


## Bioconductor setup

```{r}
library(DESeq2)

# Read in the data
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <- read.csv("airway_metadata.csv")

head(counts)
head(metadata)
```

>> Q1. How many genes are in this dataset?

```{r}
nrow(counts)
```

>> Q2. How many ‘control’ cell lines do we have? 

```{r}
table(metadata$dex)
```

## Toy differential gene expression

```{r}
# Get control samples
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[,control$id]
control.mean <- rowSums(control.counts)/4
head(control.mean)

# Get treated samples
treated <- metadata[metadata[,"dex"]=="treated",]
treated.counts <- counts[,treated$id]
treated.mean <- rowSums(treated.counts)/4
head(treated.mean)
```
>> Q3. How would you make the above code in either approach more robust? Is there a function that could help here?

Replace /4 with /ncol

>> Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

```{r}
treated <- metadata[metadata[,"dex"]=="treated",]
treated.counts <- counts[,treated$id]
treated.mean <- rowSums(treated.counts)/ncol(treated.counts)
head(treated.mean)
```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
colSums(meancounts)
```

>> Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

```{r}
plot(meancounts[,1], meancounts[,2],
     xlab="Control", ylab="Treated")
```

>> Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot?

Use `geom_point()`

>> Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this?

```{r}
plot(meancounts[,1], meancounts[,2],
     xlab="Control", ylab="Treated",
     log="xy")
```

The `log="xy"` argument sets both axes to log scale. This is useful because most genes cluster near zero and the log scale spreads them out so we can see the differences better.

```{r}
# Add log2 fold change
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)

# Remove zero count genes
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)
to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
```

>> Q7. What is the purpose of the `arr.ind` argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?

The `arr.ind=TRUE` argument makes `which()` return both the row and column positions of zero values, not just a single index. Take the first column (rows) and use `unique()` to avoid counting a gene twice if it has zeros in both samples.

```{r}
up.ind <- mycounts$log2fc > 2
sum(up.ind)
```

>> Q8. Using the `up.ind` vector above can you determine how many up regulated genes we have at the greater than 2 fc level?

250 genes are upregulated (log2fc > 2)

```{r}
down.ind <- mycounts$log2fc < (-2)
sum(down.ind)
```

>> Q9. Using the `down.ind` vector above can you determine how many down regulated genes we have at the greater than 2 fc level?

367 genes are downregulated (log2fc < -2)

>> Q10. Do you trust these results? Why or why not?

No because we only looked at fold change and not statistical testing. A gene can have a large fold change but still not be statistically significant. We need DESeq2 to properly assess this. 

## Setting up for DESeq

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts,
                              colData=metadata,
                              design=~dex)
dds
```

```{r}
dds <- DESeq(dds)
res <- results(dds)
res
```

```{r}
summary(res)
```

## Principal Component Analysis (PCA)

```{r}
vsd <- vst(dds, blind=FALSE)
plotPCA(vsd, intgroup=c("dex"))
```

```{r}
# PCA with ggplot2

library(ggplot2)

pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
head(pcaData)

percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData) +
  aes(x=PC1, y=PC2, color=dex) +
  geom_point(size=3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

## DESeq analysis

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```
Using a more strict significant cutoff of alpha=0.05, we still find many differentially expressed genes between treated and control samples.

## Adding annotation data

```{r}
library(AnnotationDbi)
library(org.Hs.eg.db)

res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     keytype="ENSEMBL",
                     column="SYMBOL",
                     multiVals="first")

res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     keytype="ENSEMBL",
                     column="ENTREZID",
                     multiVals="first")

head(res)
```

>> Q11. Run the `mapIds()` function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called `res$entrez`, `res$uniprot` and `res$genename`.

```{r}
res$uniprot <- mapIds(org.Hs.eg.db,
                      keys=row.names(res),
                      column="UNIPROT",
                      keytype="ENSEMBL",
                      multiVals="first")

res$genename <- mapIds(org.Hs.eg.db,
                       keys=row.names(res),
                       column="GENENAME",
                       keytype="ENSEMBL",
                       multiVals="first")

head(res)
```

Using `mapIds()` two more times with `column="UNIPROT"` and `column="GENENAME"` we can add UniProt accession numbers and full gene names to our results table. 

## Data Visualization

```{r}
plot(res$log2FoldChange, -log(res$padj),
     xlab="Log2(FoldChange)",
     ylab="-Log(P-value)")
```

```{r}
# Custom colored volcano plot
mycols <- rep("gray", nrow(res))
mycols[abs(res$log2FoldChange) > 2] <- "red"
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2)
mycols[inds] <- "blue"

plot(res$log2FoldChange, -log(res$padj),
     col=mycols,
     ylab="-Log(P-value)",
     xlab="Log2(FoldChange)")

abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```
The volcano plot shows fold change vs statistical significance. Blue points are genes with both large fold change and high significance. Red points have large fold change but aren't statistically significant. Grey points don't meet either threshold.







